SIMD-vectorized Chemical Source Term Evaluation

### 2017-02-21 paper.tex
vecwidth hyphens

A SIMD instruction utilizes specialized vector processing hardware to execute the same floating point operation (e.g. multiplication, division, etc.) on a several different data concurrently (Fig.~\ref{F:SIMD}); the number of concurrent operations is known as the vector-width\footnote{OpenCL allows for use of vector-widths not equal to the hardware vector-width via implicit conversion, and may provide some performance benefit as will be studied here}.
Nearly all modern CPUs have vector processing units capable of running SIMD instructions, typically capable of handling 2--4 concurrent double precision operations.
Additionally, specialized hardware accelerators, e.g. Intel's Xeon Phi co-processor (or MIC), have been developed that have tens of cores with very wide vector widths (e.g. 4--8 double precision operations); these very wide vector widths are also expected on forthcoming Intel CPUs (the Skylake Xeon and Cannon Lake architectures).

@@

\begin{figure}[htb]
  \centering
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{SIMD_SIMT_Order_comparison}
    \caption{Runtime in milliseconds per thermo-chemical state of shallow-SIMD and SIMT-vectorization on a single CPU core for both ``C'' (filled symbols) and ``F'' (empty symbols) data ordering.}
    \label{F:SIMDComp}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{SIMD_Vecwidth_comparison}
    \caption{Speedup of a ``C''-ordered shallow-SIMD vectorization on a single CPU core when varying the supplied vector-width.}
    \label{F:Veccomp}
  \end{subfigure} \\
  \begin{subfigure}[t]{0.5\textwidth}
    \includegraphics[width=\textwidth]{SIMD_scaling}
    \caption{Strong parallel scaling of a ``C''-ordered shallow-SIMD vectorization using a vector width of 16.}
    \label{F:SIMDscale}
  \end{subfigure}
\end{figure}

@@

The effect of changing the vector-width supplied to a ``C''-ordered shallow-SIMD vectorization on a single CPU core is investigated in Fig.~\ref{F:Veccomp}.
Significant speedups are seen with larger vector-widths for the smaller models, however as the model size increases data must be loaded from increasingly distant memory locations and the performance of all three vector widths becomes similar.
Finally in Fig.~\ref{F:SIMDscale} the strong parallel scaling of the chemical source term evaluation is examined for all four models with a ``C''-ordered shallow SIMD-vectorization and vector-width of 16:
Interestingly the efficiency for the hydrogen model decreases fastest, while the larger models have similar efficiencies; the isopentanol model has the highest efficiency in all cases.
This suggests that evaluation of the chemical-source terms for a single thermo-chemical state for the smaller models may not saturate the processing power of the CPU, some efficiency may be lost due to context switching within OpenCL itself.
Additionally, OpenCL allows assigning of work to devices by shared memory resources e.g., a L3 cache, which was not pursued in this study for simplicity; this should be investigated in future work to determine the effect on scaling efficiency.

\section{Conclusions}
In this work, a shallow SIMD-vectorized chemical source term evaluation OpenCL code generator was developed and validated against Cantera~\cite{Cantera} for a wide range of chemical kinetic models~\cite{Burke:2011fh,smith_gri-mech_30,Wang:2007,Sarathy:2013jr}.
Significant speedups of up to \SIrange{1.88}{2.42}{$\times$} over a baseline SIMT-vectorized code were observed.
Two data-ordering schemes were investigated, showing a clear performance benefit for use of the ``C'' (row-major) ordering for the shallow vectorized code.
Further, this study suggests that a deep SIMD-vectorized code---currently under development---may see even greater accelerations due to expected increases in cache-hit rates.
Larger vector widths were found to provide accelerations for the smaller models studied, but performance was similar over all vector widths tested for the largest models tested.
Finally, the strong parallel scaling efficiency was examined for the shallow vectorized code; the larger models exhibited higher scaling efficiency, possibly due to failure of the \ce{H2}\slash\ce{CO} model to saturate the CPU throughput.
Future extensions of this work will include development of a SIMD\slash SIMT-accelerated sparse-analytical Jacobian code targeted at CPUs, GPUs and MICs.

>>>>

A SIMD instruction utilizes specialized vector processing hardware to execute the same floating point operation (e.g. multiplication, division, etc.) on a several different data concurrently (Fig.~\ref{F:SIMD}); the number of concurrent operations is known as the vector-width\footnote{OpenCL allows for use of vector-widths not equal to the hardware vector-width via implicit conversion, and may provide some performance benefit as will be studied here}.
Nearly all modern CPUs have vector processing units capable of running SIMD instructions, typically capable of handling 2--4 concurrent double precision operations.
Additionally, specialized hardware accelerators, e.g. Intel's Xeon Phi co-processor (or MIC), have been developed that have tens of cores with very wide vector-widths (e.g. 4--8 double precision operations); these very wide vector-widths are also expected on forthcoming Intel CPUs (the Skylake Xeon and Cannon Lake architectures).

@@

\begin{figure}[htb]
  \centering
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{SIMD_SIMT_Order_comparison}
    \caption{Runtime in milliseconds per thermo-chemical state of shallow-SIMD and SIMT-vectorization on a single CPU core for both ``C'' (filled symbols) and ``F'' (empty symbols) data ordering.}
    \label{F:SIMDComp}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{SIMD_Vecwidth_comparison}
    \caption{Speedup of a ``C''-ordered shallow-SIMD vectorization on a single CPU core when varying the supplied vector-width.}
    \label{F:Veccomp}
  \end{subfigure} \\
  \begin{subfigure}[t]{0.5\textwidth}
    \includegraphics[width=\textwidth]{SIMD_scaling}
    \caption{Strong parallel scaling of a ``C''-ordered shallow-SIMD vectorization using a vector-width of 16.}
    \label{F:SIMDscale}
  \end{subfigure}
\end{figure}

@@

The effect of changing the vector-width supplied to a ``C''-ordered shallow-SIMD vectorization on a single CPU core is investigated in Fig.~\ref{F:Veccomp}.
Significant speedups are seen with larger vector-widths for the smaller models, however as the model size increases data must be loaded from increasingly distant memory locations and the performance of all three vector-widths becomes similar.
Finally in Fig.~\ref{F:SIMDscale} the strong parallel scaling of the chemical source term evaluation is examined for all four models with a ``C''-ordered shallow SIMD-vectorization and vector-width of 16:
Interestingly the efficiency for the hydrogen model decreases fastest, while the larger models have similar efficiencies; the isopentanol model has the highest efficiency in all cases.
This suggests that evaluation of the chemical-source terms for a single thermo-chemical state for the smaller models may not saturate the processing power of the CPU, some efficiency may be lost due to context switching within OpenCL itself.
Additionally, OpenCL allows assigning of work to devices by shared memory resources e.g., a L3 cache, which was not pursued in this study for simplicity; this should be investigated in future work to determine the effect on scaling efficiency.

\section{Conclusions}
In this work, a shallow SIMD-vectorized chemical source term evaluation OpenCL code generator was developed and validated against Cantera~\cite{Cantera} for a wide range of chemical kinetic models~\cite{Burke:2011fh,smith_gri-mech_30,Wang:2007,Sarathy:2013jr}.
Significant speedups of up to \SIrange{1.88}{2.42}{$\times$} over a baseline SIMT-vectorized code were observed.
Two data-ordering schemes were investigated, showing a clear performance benefit for use of the ``C'' (row-major) ordering for the shallow vectorized code.
Further, this study suggests that a deep SIMD-vectorized code---currently under development---may see even greater accelerations due to expected increases in cache-hit rates.
Larger vector-widths were found to provide accelerations for the smaller models studied, but performance was similar over all vector-widths tested for the largest models tested.
Finally, the strong parallel scaling efficiency was examined for the shallow vectorized code; the larger models exhibited higher scaling efficiency, possibly due to failure of the \ce{H2}\slash\ce{CO} model to saturate the CPU throughput.
Future extensions of this work will include development of a SIMD\slash SIMT-accelerated sparse-analytical Jacobian code targeted at CPUs, GPUs and MICs.
